---
title: "Influence of Native Sea Urchins on Biofouling in RI Oyster Farms"
author: "Christopher Jenkins"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pwr) # load packages
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)

#load data, load entire file into the R workspace
urchins_raw <- read.csv("~/R/CJ_Final_Project/urchins_example_data.csv", na.strings = c("N/A", "NA"))

#convert non-NA gis to numerics
urchins_raw[] <- apply(urchins_raw, 2, function(x) {
  numeric_vals <- as.numeric(x[!is.na(x) & !grepl("[[:alpha:]]", x)])
  replace(x, !is.na(x) & !grepl("[[:alpha:]]", x), numeric_vals)
})

#Testing
urchins_cleaned <- urchins_raw # new cleaner dataset to alter
#URCHIN DENSITY effect on BIOFOULING
urchins_cleaned$urchin_density <- factor(urchins_cleaned$urchin_density) #urchin density control variable needs to be a factor
urchins_cleaned$depth <- factor(urchins_cleaned$depth) #depth control variable needs to be a factor
model_biofouling <- aov(biofouling_end_rank ~ urchin_density*depth, data = urchins_cleaned) #anova for both biofouling and depth
#output results
summary(model_biofouling) #summary; technically only if significant results are found we continue to post hoc analysis but for this we'll be conducting it for all of them
tukey_biofouling <- TukeyHSD(model_biofouling) #post hoc analysis to identify which levels in particular are significant
#save results with significance marked
tukey_summary_biofoul <- as.data.frame(tukey_biofouling$'urchin_density') #specify what results to save
tukey_summary_biofoul$sig <- ifelse(tukey_summary_biofoul$'p adj' < 0.05, "*", " ") #add additional row that marks significant results
tukey_summary_depth <- as.data.frame(tukey_biofouling$'depth') #same for depth, check if depth has influence
tukey_summary_depth$sig <- ifelse(tukey_summary_depth$'p adj' < 0.05, "*", " ") #mark significant figures

#plot biofouling
plot_biofouling <- ggplot(urchins_cleaned, aes(x = urchin_density, y = biofouling_end_rank)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Urchin_density vs. Biofouling Ranking", x = "Urchin Density", y = "Biofouling Rank") +
  scale_y_continuous(limits = c(0, max(urchins_cleaned$biofouling_end_rank) + 1)) +
  theme_classic()

#plot depth
plot_depth <- ggplot(urchins_cleaned, aes(x = depth, y = biofouling_end_rank)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Cage Depth vs. Biofouling Ranking", x = "Cage Depth", y = "Biofouling Rank") +
  scale_y_continuous(limits = c(0, max(urchins_cleaned$biofouling_end_rank) + 1)) +
  theme_classic()

#Urchin Density and Mortality
#same as above
model_mort <- aov(mortality ~ urchin_density, data = urchins_cleaned) 
summary(model_mort) 
tukey_mort <- TukeyHSD(model_mort) 
tukey_mort <- as.data.frame(tukey_mort$'urchin_density') 
tukey_mort$sig <- ifelse(tukey_mort$'p adj' < 0.05, "*", " ") 

plot_mort <- ggplot(urchins_cleaned, aes(x = urchin_density, y = mortality)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Urchin Density vs. Urchin Mortalities", x = "Urchin Density", y = "Mortalities") +
  scale_y_continuous(limits = c(0, max(urchins_cleaned$mortality) + 1)) +
  theme_classic()

#URCHIN DENSITY effect on URCHIN QUALITY metrics
#includes GI and change in TD
#get differences in TD
# find test diameter starting average
td_start_cols <- grep("td_start", colnames(urchins_raw))
td_start_means <- rowMeans(urchins_raw[, td_start_cols], na.rm = TRUE)
urchins_cleaned$td_start_avg <- td_start_means

# find test diameter ending average
td_end_cols <- grep("td_end", colnames(urchins_raw))
td_end_means <- rowMeans(urchins_raw[, td_end_cols], na.rm = TRUE)
urchins_cleaned$td_end_avg <- td_end_means
#subtract the two columns
urchins_cleaned$td_diff <- urchins_cleaned$td_end_avg - urchins_cleaned$td_start_avg

#now get the GI %
long_urchins <- urchins_cleaned %>% # since the GI% is captured in several columns we need to reorganize the dataset so relevant information is in the same row
  gather(key = "variable", value = "gi_values", -cage_number, -urchin_density, -biofouling_end_rank, -depth, -mortality, -td_diff)
long_urchins <- na.omit(long_urchins)

#get all the GI rows with all the TD differences and conduct nested ANOVA
gi_data <- long_urchins[grep("gi", long_urchins$variable), ] # all rows that contain GI
#make td_diff a factor for analysis
gi_data$td_diff <- as.factor(gi_data$td_diff)
model_gi <- aov(gi_values ~ urchin_density, data = gi_data) #anova between GI% and urchin density
summary(model_gi) #as before, if we see significant results we move on to Tukey post-hoc
tukey_gi <- TukeyHSD(model_gi)
#save data for both GI and TD
tukey_summary_gi <- as.data.frame(tukey_gi$'urchin_density') #specify what results to save
tukey_summary_gi$sig <- ifelse(tukey_summary_gi$'p adj' < 0.05, "*", " ") #add additional row that marks significant results

gi_data$td_diff <- as.numeric(gi_data$td_diff)
model_td <- aov(td_diff ~ urchin_density, data = gi_data)
summary(model_td)
tukey_td <- TukeyHSD(model_td)
tukey_summary_td <- as.data.frame(tukey_td$'urchin_density')
tukey_summary_td$sig <- ifelse(tukey_summary_td$'p adj' < 0.05, "*", " ")

#plot for GI
plot_gi <- ggplot(gi_data, aes(x = urchin_density, y = gi_values)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Urchin Density vs. Gonad Index", x = "Urchin Density", y = "Gonad Index (%)") +
  scale_y_continuous(limits = c(0, max(gi_data$gi_values) + 1)) +
  theme_classic()

#plot for TD
gi_data$td_diff <- as.character(gi_data$td_diff)
gi_data$td_diff <- as.numeric(gi_data$td_diff)
plot_td <- ggplot(gi_data, aes(x = urchin_density, y = td_diff)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Urchin Density vs. Change in Test Diameter", x = "Urchin Density", y = "Change in Test Diameter (mm)") +
  scale_y_continuous(limits = c(0, max(gi_data$td_diff) + 1)) +
  theme_classic()

#URCHIN DENSITY effect on OYSTER QUALITY - change in SL
# find oyster shell length starting average
shell_length_start <- grep("shell_length_start", colnames(urchins_raw))
shell_length_start_means <- rowMeans(urchins_raw[, shell_length_start], na.rm = TRUE)
urchins_cleaned$shell_length_start_avg <- shell_length_start_means
# find oyster shell length ending average
shell_length_end <- grep("shell_length_end", colnames(urchins_raw))
shell_length_end_means <- rowMeans(urchins_raw[, shell_length_end], na.rm = TRUE)
urchins_cleaned$shell_length_end_avg <- shell_length_end_means
#subtract the two columns
urchins_cleaned$shell_length_diff <- urchins_cleaned$shell_length_end_avg - urchins_cleaned$shell_length_start_avg

#perform anova between differences in SL and urchin density
urchins_cleaned$biofouling_end_rank <- factor(urchins_cleaned$biofouling_end_rank)
model_sl <- aov(shell_length_diff ~ urchin_density, data = urchins_cleaned)
#output results
summary(model_sl)
tukey_slu <- TukeyHSD(model_sl)
#save data for both GI and TD
tukey_summary_slu <- as.data.frame(tukey_slu$'urchin_density') #specify what results to save
tukey_summary_slu$sig <- ifelse(tukey_summary_slu$'p adj' < 0.05, "*", " ") #add additional row that marks significant results

#plot of urchin density vs shell length growth
plot_slu <- ggplot(urchins_cleaned, aes(x = urchin_density, y = shell_length_diff)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Urchin Density vs. Change in Oyster Shell Length", x = "Urchin Density", y = "Change in Oyster Shell Length (mm)") +
  scale_y_continuous(limits = c(0, max(urchins_cleaned$td_diff) + 1)) +
  theme_classic()

#BIOFOULING effect on OYSTER QUALITY
urchins_cleaned$biofouling_end_rank <- as.factor(urchins_cleaned$biofouling_end_rank)
model_slb <- aov(shell_length_diff ~ biofouling_end_rank, data = urchins_cleaned)
#output results
summary(model_slb)
tukey_slb <- TukeyHSD(model_slb)
#save data for both GI and TD
tukey_summary_slb <- as.data.frame(tukey_slb$'biofouling_end_rank') #specify what results to save
tukey_summary_slb$sig <- ifelse(tukey_summary_slb$'p adj' < 0.05, "*", " ") #add additional row that marks significant results

plot_slb <- ggplot(urchins_cleaned, aes(x = biofouling_end_rank, y = shell_length_diff)) +
  geom_violin(fill = "lightblue") + # violin plot
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title = "Biofouling Rank vs. Change in Oyster Shell Length", x = "Biofouling Rank", y = "Change in Oyster Shell Length (mm)") +
  scale_y_continuous(limits = c(0, max(urchins_cleaned$shell_length_diff) + 1)) +
  theme_classic()
```

## Introduction

This analysis requires a dataset in long form, arranged such that treatment cages are in rows and data collected for each row is found in the columns. Data collected includes: mortality, biofouling end rank, test diameter at start of experiment and test diameter at end of experiment, shell length at start of experiment and shell length at end of experiment, and GI%. This dataset assumes GI% calculations have already been performed on the raw data.

## Methods

In order to prove our assumptions that determine our decision to use an ANOVA test, we can check that our response variables have approximately similar variances using an F-test for equality of variances at P>0.05, where if P>0.05 we can assume the variances are similar.
```{r, echo=FALSE}
avg_gi <- aggregate(gi_values ~ cage_number, data = gi_data, FUN = mean) #using average of GI to compare to average test diameter difference
urchins_cleaned <- merge(urchins_cleaned, avg_gi, by = "cage_number", all.x = TRUE)
#values must be numerics for variance test
urchins_cleaned$biofouling_end_rank <- as.numeric(urchins_cleaned$biofouling_end_rank)
urchins_cleaned$gi_values <- as.numeric(urchins_cleaned$gi_values)
urchins_cleaned$shell_length_diff <- as.numeric(urchins_cleaned$shell_length_diff)
#variance tests
var.test(urchins_cleaned$biofouling_end_rank, urchins_cleaned$shell_length_diff)
var.test(urchins_cleaned$biofouling_end_rank, urchins_cleaned$gi_values)
var.test(urchins_cleaned$shell_length_diff, urchins_cleaned$gi_values)
```
If variances are unequal, we can use an alternative test method such as Welch's ANOVA or Kruskal-Wallis test.

Additionally, ANOVA assumes that all response variables are independent of one another. This can be tested using a pairwise correlation. We will be using a Kendall coefficient of correlation as the sample size is small and the biofouling is ordinal in nature. Below are the results between biofouling rank and oyster shell growth rate:
```{r, echo=FALSE}
#insert pairwise correlation here with cor() function
urchins_cleaned$biofouling_end_rank <- as.numeric(urchins_cleaned$biofouling_end_rank)
cor(urchins_cleaned$biofouling_end_rank, urchins_cleaned$shell_length_diff, method = c("kendall"))
```

Same test on GI% and test diameter growth rate:
```{r, echo=FALSE}
#changes required so that the correlation test can work due to N/A and NaN values when urchin density = 0
cor_test <- subset(urchins_cleaned, select = c("gi_values", "td_diff"))
cor_test <- na.omit(cor_test)
cor(cor_test$td_diff, cor_test$gi_values, method = c("kendall"))
```

Same test on GI% and biofouling rank
```{r, echo=FALSE}
cor_test2 <- subset(urchins_cleaned, select = c("gi_values", "biofouling_end_rank"))
cor_test2 <- na.omit(cor_test2)
cor(cor_test2$biofouling_end_rank, cor_test2$gi_values, method = c("kendall"))
```

If correlations are found, this can be used to reinforce our understanding of how urchin presence/density influences oyster quality and urchin quality. We can focus on more specific variables with stronger effects. Non-correlation means ANOVAs should be performed on all variables.

## Analysis

Here we can view the analysis of our data using one-way ANOVAs. Each section produces a violin plot of the data with standard error bars and median highlighted, as well as a summary of a tukey's honest significant differences post-hoc analysis with statistically significant differences highlighted. This analysis investigates the relationship between the independent variable urchin density and biofouling rank, mortality, test diameter, GI%, and oyster shell length. We also check the covariate for influence on biofouling, as well as an additional test between biofouling rank and oyster shell length where we assume biofouling rank to be an independent variable.

Influence of cage depth on biofouling ranking
```{r, echo=FALSE}
plot(plot_depth)
knitr::kable(tukey_summary_depth, format = "markdown")
```
This analysis checks the covariate depth for influence on biofouling ranking. For this example, there is no influence of cage depth on biofouling ranking and we can therefore safely assume that all the cages are functionally replicates of one another between depths (for a given urchin density); if we do see differences in biofouling ranking this flags the user to perform a deeper analysis between urchin density AND depth on the response variables. This would be performed using an ANCOVA test (Analysis of Covariance) using the lm() function.

Influence of urchin density on biofouling rankings:
```{r, echo=FALSE}
plot(plot_biofouling)
knitr::kable(tukey_summary_biofoul, format = "markdown")
```

Influence of urchin density on mortality rate:
```{r, echo=FALSE}
plot(plot_mort)
knitr::kable(tukey_mort, format = "markdown")
```

Influence of urchin density on urchin quality metric test diameter:
```{r, echo=FALSE}
plot(plot_td)
knitr::kable(tukey_summary_td, format = "markdown")
```

Influence of urchin density on quality metric Gonad Index (%):
```{r, echo=FALSE}
plot(plot_gi)
knitr::kable(tukey_summary_gi, format = "markdown")
```

Influence of urchin density on the oyster quality metric shell length:
```{r, echo=FALSE}
plot(plot_slu)
knitr::kable(tukey_summary_slu, format = "markdown")
```

Influence of biofouling on the oyster quality metric shell length:
```{r, echo=FALSE}
plot(plot_slb)
knitr::kable(tukey_summary_slb, format = "markdown")
```

Power analysis:
```{r, echo=FALSE}
pwr.t.test(d = 0.25, sig.level = 0.05, power = 0.8)$n
pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.8)$n
pwr.t.test(d = 0.75, sig.level = 0.05, power = 0.8)$n
```
This power analysis is produced for small, medium, and large differences in means. This is useful as a reference in determining whether or not the sampling size plan we're using is large enough to appropriately account for Type II error.